The Design and Implementation of the Synchronous Language \CEU

### 2016-03-31 paper.tex
...

% TODO: SCCharts?, SynchCharts?

@@

% TODO: par/or vs abort, strong/weak abortion

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
%We finish with a summary of our design
    %(\ref{sec.ceu.discussion}).

@@

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO: Such analsysis is orthogonal

@@

4. WORST CASE REACTION TIME ANALYSIS
TODO: requires this, what about this in the context of MThreads??
Towards Direct Execution of Esterel Programs on Reactive Processors
3.1 Preserving the Synchrony Hypothesis
The synchronous model implies that reactions to input signals are instantaneous 
and occur at discrete logical instants called ticks.
As a result, a sequence of actions may need to be performed within the duration 
of a single tick.
In order to model the logical tick of Esterel in RePIC, a tick with variable 
length in terms of absolute time is implemented.
The duration of one tick is determined by the time required to execute all the 
instantaneous instructions that are placed between any two consecutive await 
(i.e. TAWAIT, SAWAIT or CAWAIT) or halt instructions which are the tick 
delimiting instructions.
(TODO: tem figuras)
\end{comment}

@@

% TODO: .. will probably expect that duplicating the  emit inc => &v; line 
% would result in v==3.

@@

Note that the illustrative example of Figure~\ref{lst.fin} does not manipulate 
pointers (i.e., the resource is a \emph{singleton}).
That case is an example of a bad and unsafe API to expose to \CEU because the 
compiler will not enforce the use of finalization.
%
%TODO: @nohold
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

First-class timers simplify the implementation of application-defined 
simultaneity.
%
Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
\code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Note also that the application in \CEU could define an external input 
\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
10) to recover the abstract notion of simultaneity.

@@

@@

>>>>

% TODO-no: SCCharts?, SynchCharts?

@@

% TODO-no: par/or vs abort, strong/weak abortion

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
%We finish with a summary of our design
    %(\ref{sec.ceu.discussion}).
%
We provide the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction of a given event.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO-no: Such analsysis is orthogonal

@@

4. WORST CASE REACTION TIME ANALYSIS
TODO-no: requires this, what about this in the context of MThreads??
Towards Direct Execution of Esterel Programs on Reactive Processors
3.1 Preserving the Synchrony Hypothesis
The synchronous model implies that reactions to input signals are instantaneous 
and occur at discrete logical instants called ticks.
As a result, a sequence of actions may need to be performed within the duration 
of a single tick.
In order to model the logical tick of Esterel in RePIC, a tick with variable 
length in terms of absolute time is implemented.
The duration of one tick is determined by the time required to execute all the 
instantaneous instructions that are placed between any two consecutive await 
(i.e. TAWAIT, SAWAIT or CAWAIT) or halt instructions which are the tick 
delimiting instructions.
(TODO-no: tem figuras)
\end{comment}

@@

% TODO-no: .. will probably expect that duplicating the  emit inc => &v; line 
% would result in v==3.

@@

Note that the illustrative example of Figure~\ref{lst.fin} does not manipulate 
pointers (i.e., the resource is a \emph{singleton}).
That case is an example of a bad and unsafe API to expose to \CEU because the 
compiler will not enforce the use of finalization.
%
%TODO-no: @nohold
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

First-class timers simplify the implementation of application-defined 
simultaneity.
%
Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
\code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Conversely, an environment for \CEU could provide \code{TICK} input and 
broadcast it periodically to allow applications to recover the abstract notion 
of simultaneity (e.g., changing ln. 10 to \code{await TICK}).
%the expiring event from \code{200ms} to \code{TICK} in ln.  10).

@@

In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
in comparison to hand-crafted event-driven code in \emph{nesC}~\cite{wsn.nesc} 
(a \C variant) available in a stable codebase%
\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}.
%
Figure~\ref{fig.eval} compares source size (\emph{tokens}), binary size 
(\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
network protocols and a radio driver.
The small increase in resource usage shows that \CEU fits the context of 
constrained embedded systems which is typically reserved to \C only.

\begin{figure*}[t]
\includegraphics[width=\textwidth,clip=true,trim=-80px 385px 300px 40px]{eval}
\caption{ Resource usage for \CEU and \emph{nesC} in the domain of sensor 
networks.
\label{fig.eval}
}
\end{figure*}

@@

- TODO: no available Esterel compiler for the constrained target platform
    - we compare with C, which is a benchmark, technically unbeatable for 
      resource efficiency
